### 2024-03-01 manuscript.tex
Addressed reviewer comments about clarifying introduction of contrasts algorithm
Added reference to contrasts figure in supporting.tex
Added citations to Felsenstein's algorithm in main text
Use disorder scores consistently instead of only scores

\subsection{Intrinsic disorder is poorly conserved in some proteins}
Though the substitution models reveal specific patterns of evolution at the level of individual residues, the large amounts of sequence divergence between many orthologous IDRs implies their evolution is not well-described by fine-scale models of residue substitution. Given the growing evidence that IDRs are constrained to conserve distributed properties, we instead turned towards characterizing their evolution in terms of 82 disorder-associated ``molecular features'' obtained from the previous study of IDRs in the yeast proteome. However, before conducting an in-depth analysis of these features, we examined the disorder score traces in greater detail and were struck by the significant variability between species. For each residue in the input sequence, AUCPreD returns a score between 0 and 1 where higher values indicate higher confidence in a prediction of intrinsic disorder. In some alignments, the scores vary by nearly this entire range at a given position even when there is a relatively high level of sequence identity (Fig~\ref{fig:scores}A). Though we cannot fully eliminate the possibility that this result is a prediction artifact, the strong performance of AUCPreD in a recent assessment of disorder predictors suggests the observed variability reflects true changes in these residues' propensity for disorder~\cite{Necci2021}.

To better understand the relationship of this variability to differences in the regions' biophysical properties, we sought to correlate the average disorder score of the segments in a region with their molecular features. However, as the sequences are not independent but instead related by a hierarchical structure which reflects their evolutionary relationships, any features derived from them are unsuitable for direct use in many standard statistical procedures. In the most severe cases, traits derived from clades of closely-related species can effectively act as duplicate observations, which can yield spurious correlations. We therefore applied the method of contrasts to both the scores and the features. This algorithm takes differences between adjacent nodes in the phylogenetic tree relating the species to generate ``contrasts,'' which, under some general assumptions of the underlying evolutionary process, are independent and identically-distributed and therefore appropriate for use in correlation analyses. The resulting feature contrasts have varying degrees of correlation with the score contrasts (Fig~\ref{fig:scores}B-C). Some, like isopoint, are uncorrelated, but most are significantly, if weakly, correlated. In general, the strongest correlations are observed for features which have a direct biophysical relationship to the presence or absence of disorder, such as disorder\_fraction or hydrophobicity. Interestingly, the correlations with many motifs were statistically significant, though small in magnitude relative to the non-motif features. However, a more detailed analysis of this observation is presented in the discussion. To determine if regions with rapidly evolving disorder scores are associated with particular functions, processes, or compartments, we then extracted the regions in the upper decile of the rate distribution and performed a term enrichment analysis on their associated annotations (\ref{sfig:scores_histogram}~Fig). The most significant terms are generally related to DNA repair or extracellular structure, which suggests these processes and components are enriched in proteins whose structural state is rapidly evolving (Fig~\ref{fig:scores}E).

>>>>

\subsection{Intrinsic disorder is poorly conserved in some proteins}
Though the substitution models reveal specific patterns of evolution at the level of individual residues, the large amounts of sequence divergence between many orthologous IDRs implies their evolution is not well-described by fine-scale models of residue substitution. Given the growing evidence that IDRs are constrained to conserve distributed properties, we instead turned towards characterizing their evolution in terms of 82 disorder-associated ``molecular features'' obtained from the previous study of IDRs in the yeast proteome. However, before conducting an in-depth analysis of these features, we examined the disorder score traces in greater detail and were struck by the significant variability between species. For each residue in the input sequence, AUCPreD returns a score between 0 and 1 where higher values indicate higher confidence in a prediction of intrinsic disorder. In some alignments, the disorder scores vary by nearly this entire range at a given position even when there is a relatively high level of sequence identity (Fig~\ref{fig:scores}A). Though we cannot fully eliminate the possibility that this result is a prediction artifact, the strong performance of AUCPreD in a recent assessment of disorder predictors suggests the observed variability reflects true changes in these residues' propensity for disorder~\cite{Necci2021}.

To better understand the relationship of this variability to differences in the regions' biophysical properties, we sought to correlate the average disorder score of the segments in a region with their molecular features. However, as the sequences are not independent but instead related by a hierarchical structure which reflects their evolutionary relationships, any features derived from them are unsuitable for direct use in many standard statistical procedures. In the most severe cases, traits derived from clades of closely-related species can effectively act as duplicate observations, which can yield spurious correlations. We therefore applied the method of contrasts to both the disorder scores and the features~\cite{Felsenstein1973, Felsenstein1985}. This algorithm takes differences between adjacent nodes in the phylogenetic tree relating the species to generate ``contrasts,'' which, under some general assumptions of the underlying evolutionary process, are independent and identically-distributed and therefore appropriate for use in correlation analyses. (A sample calculation using this algorithm is shown in Fig~\ref{sfig:contrasts}.) The resulting feature contrasts have varying degrees of correlation with the score contrasts (Fig~\ref{fig:scores}B-C). Some, like isopoint, are uncorrelated, but most are significantly, if weakly, correlated. In general, the strongest correlations are observed for features which have a direct biophysical relationship to the presence or absence of disorder, such as disorder\_fraction or hydrophobicity. Interestingly, the correlations with many motifs were statistically significant, though small in magnitude relative to the non-motif features. However, a more detailed analysis of this observation is presented in the discussion. To determine if regions with rapidly evolving disorder scores are associated with particular functions, processes, or compartments, we then extracted the regions in the upper decile of the rate distribution and performed a term enrichment analysis on their associated annotations (\ref{sfig:scores_histogram}~Fig). The most significant terms are generally related to DNA repair or extracellular structure, which suggests these processes and components are enriched in proteins whose structural state is rapidly evolving (Fig~\ref{fig:scores}E).
